Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-07-17 paper.Rnw
'nother day, 'nother dollar, Donegan

\ants is a diffeomorphic registration algorithm which provides great flexibility
over the choice of transformation model, objective function, and the consistency
of the final transformation. The transformation is estimated in a hierarchical
fashion where the MRI data is subsampled, allowing large deformations to be
estimated and successively refined at later hierarchical stages (where the data
is subsampled to a finer grid). The deformation field and the objective function
are regularized with a Gaussian kernel at each level of the hierarchy. The \ants
algorithm is freely available \url{http://www.picsl.upenn.edu/ANTS/}. We used an
implementation of the \ants algorithm compatible with the MINC data format,
mincANTS \url{https://github.com/vfonov/mincANTS}.

@@

In general, across hippocampal subregions the percent error in volume of
\mb segmentations compares
favourably to the error resulting from image resampling (Figure
\ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4, and Dentate
subregions all show near or smaller percent errors. The Sibiculum and CA2/CA3
subregions show distinctly larger error than is found through resampling.  
\todo{note direction of change}.

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/subfield-montage.pdf}
  \end{centering}
  \caption{Sagittal slices from two subjects showing a comparison of the
  original Winterburn atlas subfield segmentations (at 0.3mm3-isotropic voxel
  resolution), the subsampled Winterburn segmentations (at 0.9mm3-isotropic
  voxel resolution), and the MAGeT brain labels on the subsampled atlas image.}
  \label{fig:subfield-montage}
  \todo{MAGeT on BRAVO.  Also, show coronal views}
\end{figure}

@@

\begin{verbatim}

In this manuscript we have presented the implementation and validation of the MAGeT Brain framework; a methodology that requires very few input atlases in order to provide accurate and reliable segmentations.  We have demonstrated that our methodology robustly provides accurate and consistent segmentations in populations with different ageing and neuropsychiatric characteristics.  Further, we have demonstrated that algorithmic performance is not dependent on a single definition of the hippocampus and can easily differing hippocampal definitions \cite{Winterburn,Pruessner,and SNT}.  Finally, we also demonstrate that MAGeT Brain provides accurate automatic identification of the hippocampal subfields despite contrast and resolution limitations in standard T1-weighted image volumes.

Throughout the cross validation in Experiment #1 (10-fold cross validation in the ADNI sample) we demonstrate that two parameter choices increase segmentation accuracy; namely increasing the number of atlases and the number of templates.  However after setting the parameters to 5 atlases and 15 templates that there are diminishing returns with respect to this improvement.  We were somewhat surprised to find that cross-correlation and normalized mutual information based weighted voting did little to improve segmentation accuracy.  This suggests that using an intermediate template library generated by accurate nonlinear registration methods such as ANTs \cite{Avants|} sufficiently filters sources of error such as nonlinear registration error and partial volume effects that arise through the use of nearest neighbour resampling. From our previous work on the MAGeT Brain algorithm we have shown that these increases in error are not simply a smoothing or averaging effect \cite{Chakravarty}.  Although, the goal of this manuscript was to not exhaustively test or validate multiple different voting strategies in the context of our segmentation algorithm, it is important to note that other strategies for voting are available.  For example, other groups have used the STAPLE algorithm \cite{Warfield} (or variants of the STAPLE algorithm \cite{stuff from the Duchesne group here}).  Can then be weighted each segmentation based upon its estimated performance level with respect to the other available candidate segmentations.  Further, the sensitivity and specificity parameters could also be tuned to potentially improve segmentation accuracy and reliability.  It is rather likely that using more sophisticated voting methods would have a positive effect on the overall segmentation performance.  To this end, more work is required in order to determine the source of the slight decrease in segmentation performance when the number of templates are set to an even number.  Our initial concern was that this dip in performance was a by-product of the MAGeT Brain algorithm itself;  however we determined that this pattern was also true in the analysis of multi-atlas segmentations we used in our experiments.  We believe that our voting methodology is biased towards labels with the lowest numeric values when breaking ties, thus causing the slight bias observed when using an even number of templates.  This is another area where the voting scheme could be used to improve performance.  However, it is worth noting that this limitation was previously identified by Heckemann et al. \cite{Heckemann2006a} and, subsequently, other groups have not even considered the case where an even number of candidate labels are considered (eg: \cite{Leung2010}).

Another concern is the moderate-to-small improvement observed in MAGeT Brain in comparison to multi-atlas segmentation when using the same number of atlases.  The actual benefit in using MAGeT Brain is consistency of the labelling regardless of atlas or template choice.  This is an important consideration that few have touched on previously.  The 10-fold Monte Carlo cross validation that we present in Experiment 1 is amongst one of the most stringent performed in the multi-atlas/segmentation literature.  To the best of our knowledge other groups have only ever performed a 3-fold validation \cite{????}.  This suggests that our results are reflective of a true average over the choice of parameter settings and are independent of atlas or template choice (provided the input atlases are properly segmented).

In comparison to other methodologies in the field, it is important to note that MAGeT Brain performs quite well.  Table 6 gives a survey of some of the most recent multi-atlas implementations in the fields and the kappa values that are reported that validate their results using the ADNI dataset, many of which use the same SNT labels for populating their template library and for gold standards for evaluation.  While it is difficult to compare segmentation results across studies, gold-standards, evaluation metrics, and algorithms it is worth noting that the methods summarized in Table 6 require more atlases (between 16-55) than our MAGeT Brain implementation with the Winterburn atlases \cite{Winterburn}.  Amongst these methods only 4 yields mean Kappas that are higher than the ones that are reported here.  There are some important differences between our method and these specific methods.  Others have reported the difficulty with misregistrations in candidate segmentation (ie: segmentations generated that are then input in the voxel-voting procedure \cite{Collins}.  The work of Leung \cite{Leung2010} (work also reports some of the highest kappas for the segmentation of ADNI data) tackles this problems by using an intensity threshold that is estimated heuristically at the time of segmentation.  While this works for the ADNI dataset (which is partially homogenized with respect to image acquisition and pre-processing), it is unclear if this type of heuristic is applicable to other datasets.  In all cases, these methods require more atlases than our implementation with the Winterburn atlases.  Lotjonen et al. \cite{Lotjonen} achieve highly accurate segmentation but correct their segmentations using classifications derived using an expectation maximization framework.  In their initial work, Chupin and colleagues develop their probabilistic methodology using a cohort of 8 healthy controls and 15 epilepsy patients.  In subsequent methods, they retuned their methodology using the ADNI sample with a hierarchical experimentation protocol that involved.  These methods suggest that some post-processing of the final segmentations would improve accuracy of the segmentation.  While that may be true, there is little consensus regarding how to achieve this.  Further, to the best of our knowledge, no other groups have validated their work using multiple atlas labels, different acquisitions, and disease populations in order to demonstrate the robustness of their technique.  This is one of the clear strengths of this work.  Further, unlike some of the algorithms mentioned our implementation does not require retuning for new populations or datasets as it inherently models the variability of the dataset through the template library.  However it should be noted that the increased accuracy that follows increasing the number of atlases and templates comes at an increased computational cost (O(log(n), as previously mentioned in other work \cite{Heckemann}).

In comparison to the methods that we compared and were available through the ADNI database (FreeSufer, MAPER) and the method that initialized ourselves (FSL-FIRST) we find extremely variable performance of all methods.  With the exception of FIRST all methods correlate well with the SNT volumes provided in the ADNI database.  However, FreeSufer and FIRST provide radically different definitions of the size of the hippocampus in comparison to the other methods.  Further, when estimating bias of these methods relative to SNT hippocampal volumes we see that large hippocampi are over estimated while small hippocampi are under estimated.  By comparison, our method and MAPER are far more conservative suggesting that these methods may be better suited for estimating true-positives.  While this comparison with these methods using only volume, more work is needed to better understand the differences between our method and other methods.

Finally, we have also demonstrated that our algorithmic framework is appropriate for the segmentation of hippocampal subfields in standard T1-weighted data.  This has started to be come a burgeoning topic in the segmentation literature, although very few methods are available for the segmentation of 3T data \cite{Yushkevich and van Leemput}.  However, the FreeSufer implementation of subfield segmentation.  While recent work demonstrates that subfield segmentations can be used for classification of AD, MCI, and NC, there has been no explicit validation of the methodology based on accruacy or precision. Although the initial hippocampal subfield segmentations from the Yushkevich group have been demonstrated to work on the ADNI population, there has also been no validation of their work.  In addition, their work requires some manual initialization to properly function.  Our work demonstrates that we can reliably identify the CA1, subiculum, and CA4 dentate with only modest amounts of error.  The fact that CA2/CA3 and molecular layers cannot be reasonably identified should not be surprising.  These are extremely thin and spatially convoluted regions that originally required high-resolution MRI for identification.  It is likely that they would be that extents of these regions are well below the resolution offered by standard T1-weighted images.  In fact many manual segmentation methodologies do not attempt to parse these regions either \cite{Wisse, Mueller}

@@

Experiment 1: 
- address the absence of weighted-voting effects. 
  - hypothesis: the benefits of weighted voting only outweigh the resampling
    error effects when choosing from a large library.  In this experiment, the 
    template library (during weighted voting) consists of only 20 templates, so
    the most we could hope for is a "squashing" upwards of the curve towards 
    the 20-template limit (which is the same across all cases in this experiment).
  - we do see a slight "squashing" effect, with XCORR more strongly than NMI. 
    Perhaps in future experiments with larger template libraries, this could be
    explored further.
- Why don't we use STAPLE? What would we expect from STAPLE or other fusion methods? 
  - couldn't get STAPLE to work with our image formats
  - Idea: expect smoothness across range of templates (no even # dips, below)
  - perhaps more sophisticated fusion methods would boost results over majority 
    vote based techniques.

>>>>

\ants is a diffeomorphic registration algorithm which provides great flexibility
over the choice of transformation model, objective function, and the consistency
of the final transformation \citep{Avants}. The transformation is estimated in a
hierarchical fashion where the MRI data is subsampled, allowing large
deformations to be estimated and successively refined at later hierarchical
stages (where the data is subsampled to a finer grid). The deformation field and
the objective function are regularized with a Gaussian kernel at each level of
the hierarchy. The \ants algorithm is freely available
\url{http://www.picsl.upenn.edu/ANTS/}. We used an implementation of the \ants
algorithm compatible with the MINC data format, mincANTS
\url{https://github.com/vfonov/mincANTS}.

@@

In general, across hippocampal subregions the percent error in volume of
\mb segmentations compares
favourably to the error resulting from image resampling (Figure
\ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4, and Dentate
subregions all show near or smaller percent errors. The Sibiculum and CA2/CA3
subregions show distinctly larger error than is found through resampling.  
\todo{note direction of change.}

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/subfield-montage.pdf}
  \end{centering}
  \caption{Sagittal slices from two subjects showing a comparison of the
  original Winterburn atlas subfield segmentations (at 0.3mm3-isotropic voxel
  resolution), the subsampled Winterburn segmentations (at 0.9mm3-isotropic
  voxel resolution), and the \mb labels on the subsampled atlas image.}
  \label{fig:subfield-montage}
  \todo{MAGeT on BRAVO.  Also, show coronal views}
\end{figure}

@@

In this manuscript we have presented the implementation and validation of the
\mb framework -- a methodology that requires very few input atlases in
order to provide accurate and reliable segmentations.  We have demonstrated that
our methodology robustly provides accurate and consistent segmentations in
populations with different ageing and neuropsychiatric characteristics.
Further, we have demonstrated that algorithmic performance is not dependent on a
single definition of the hippocampus but is effective with differing hippocampal
definitions \citep{Winterburn2013,Pruessner2000,Hsu2002}.  Finally, we also demonstrate
that \mb provides accurate automatic identification of the hippocampal
subfields despite contrast and resolution limitations in standard T1-weighted
image volumes.

Throughout the cross validation in Experiment 1 (10-fold cross validation in the
\adnidataset dataset subsample) we find that two parameter choices improve 
segmentation accuracy: increasing the number of atlases, and the number of
templates. However, after setting the parameters to 5 atlases and 15 templates
there are diminishing returns with respect to this improvement. We were
somewhat surprised to find that cross-correlation and normalized mutual
information based weighted voting did little to improve segmentation accuracy.
This suggests that using an intermediate template library generated by accurate
nonlinear registration methods such as \ants \cite{Avants} sufficiently filters
sources of error such as nonlinear registration error and partial volume effects
that arise through the use of nearest neighbour resampling. From our previous
work on the \mb algorithm we have shown that these increases in error
are not simply a smoothing or averaging effect \cite{Chakravarty2013}.

Although, the goal of this manuscript was to not exhaustively test or validate
multiple different voting strategies in the context of our segmentation
algorithm, it is important to note that other strategies for voting are
available.  For example, other groups have used the STAPLE algorithm
\cite{Warfield2004} (or variants of the STAPLE algorithm \cite{stuff from the
Duchesne group here} which weights each segmentation based upon its estimated
performance level with respect to the other available candidate segmentations.
Further, the sensitivity and specificity parameters could also be tuned to
potentially improve segmentation accuracy and reliability.  It is rather likely
that using more sophisticated voting methods would have a positive effect on the
overall segmentation performance.

To this end, more work is required in order to determine the source of the
slight decrease in segmentation performance when the number of templates are set
to an even number.  Our initial concern was that this dip in performance was a
by-product of the \mb algorithm itself.  However, we determined that
this pattern was also true in the analysis of multi-atlas segmentations we used
in our experiments.  We believe that our majority voting methodology is biased
towards labels with the lowest numeric values when breaking ties, thus causing
the slight bias observed when using an even number of templates. This is another
area where the voting scheme could be used to improve performance. However, it
is worth noting that this limitation was previously identified by Heckemann et
al. \cite{Heckemann2006a} and, subsequently, other groups have not even
considered the potential pitfalls of an even number of candidate labels (eg.
\cite{Leung2010}).

Another concern is the moderate-to-small improvement observed in \mb in
comparison to multi-atlas segmentation when using the same number of atlases.
The actual benefit in using \mb is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo cross validation that we present
in Experiment 1 is amongst one of the most stringent performed in the
multi-atlas/segmentation literature.  To the best of our knowledge other groups
have only ever performed a 3-fold validation \cite{????}.  This suggests that
our results are reflective of a true average over the choice of parameter
settings and are independent of atlas or template choice (provided the input
atlases are properly segmented).

\todo{additionally, we could be seeing a ceiling effect: reaching the limit of
what is achievable with a multi-atlas-based segmentation method}

In comparison to other methodologies in the field, it is important to note that
\mb performs quite well.  Table 6 gives a survey of some of the most recent
multi-atlas implementations with the reported kappa values used to validate the
algorithm on the ADNI dataset (many of which use the same SNT labels for
populating their template library and as gold standards for evaluation).  While
it is difficult to compare segmentation results across studies, gold-standards,
evaluation metrics, and algorithms it is worth noting that the methods
summarized in Table 6 require more atlases (between 16-55) than our \mb
implementation with the Winterburn atlases \cite{Winterburn2013}.  Amongst these
methods only method 4 yields mean Kappas that are higher than the ones that are
reported here.

There are some important differences between our method
and these specific methods.  Others have reported the difficulty with
misregistrations in candidate segmentation (i.e. segmentations generated that are
then input in the voxel-voting procedure \cite{Collins2010}). The work of Leung
\cite{Leung2010} (work also reports some of the highest kappas for the
segmentation of ADNI data) tackles this problems by using an intensity threshold
that is estimated heuristically at the time of segmentation.  While this works
for the ADNI dataset (which is partially homogenized with respect to image
acquisition and pre-processing), it is unclear if this type of heuristic is
applicable to other datasets.  In all cases, these methods require more atlases
than our implementation with the Winterburn atlases.  Lotjonen et al.
\cite{Lotjonen2010} achieve highly accurate segmentation but correct their
segmentations using classifications derived using an expectation maximization
framework.  In their initial work, Chupin and colleagues develop their
probabilistic methodology using a cohort of 8 healthy controls and 15 epilepsy
patients.  In subsequent methods, they retuned their methodology using the ADNI
sample with a hierarchical experimentation protocol that involved.  These
methods suggest that some post-processing of the final segmentations would
improve accuracy of the segmentation.  While that may be true, there is little
consensus regarding how to achieve this.

To the best of our knowledge, no other groups have validated their work using
multiple atlas segmentation protocols, different acquisitions, and disease
populations in order to demonstrate the robustness of their technique.  This is
one of the clear strengths of this work.  Furthermore, unlike some of the
algorithms mentioned, our implementation does not require retuning for new
populations or datasets as it inherently models the variability of the dataset
through the template library.  However it should be noted that the increased
accuracy that follows increasing the number of atlases and templates comes at an
increased computational cost ($\Omicron(log(n))$), as previously mentioned in
other work \cite{Heckemann}).

In comparison to the methods that we compared and were available through the
ADNI database (FreeSufer, MAPER) and the method that we initialized ourselves
(FSL-FIRST) we find extremely variable performance of all methods.  With the
exception of FSL all methods correlate well with the SNT volumes provided in
the ADNI database.  However, FreeSufer and FIRST provide radically different
definitions of the size of the hippocampus in comparison to the other methods.
Further, when estimating bias of these methods relative to SNT hippocampal
volumes we see that large hippocampi are over estimated while small hippocampi
are under estimated.  By comparison, our method and MAPER are far more
conservative suggesting that these methods may be better suited for estimating
true-positives.  While this comparison with these methods using only volume,
more work is needed to better understand the differences between our method and
other methods.

Finally, we have also demonstrated that our algorithmic framework is appropriate
for the segmentation of hippocampal subfields in standard T1-weighted data.
This has started to become a burgeoning topic in the segmentation literature,
although very few methods are available for the segmentation of 3T data
\cite{Yushkevich2009, and VanLeemput2009}.  However, the FreeSufer implementation of
subfield segmentation.  While recent work demonstrates that subfield
segmentations can be used for classification of AD, MCI, and NC, there has been
no explicit validation of the methodology based on accruacy or precision.
Although the initial hippocampal subfield segmentations from the Yushkevich
group have been demonstrated to work on the ADNI population, there has also been
no validation of their work.  In addition, their work requires some manual
initialization to properly function.  Our work demonstrates that we can reliably
identify the CA1, subiculum, and CA4 dentate with only modest amounts of error.
The fact that CA2/CA3 and molecular layers cannot be reasonably identified
should not be surprising.  These are extremely thin and spatially convoluted
regions that originally required high-resolution MRI for identification.  It is
likely that they would be that extents of these regions are well below the
resolution offered by standard T1-weighted images.  In fact many manual
segmentation methodologies do not attempt to parse these regions either
\cite{Wisse2012, Mueller2009}

@@

\begin{verbatim}
Experiment 1: 
- address the absence of weighted-voting effects. 
  - hypothesis: the benefits of weighted voting only outweigh the resampling
    error effects when choosing from a large library.  In this experiment, the 
    template library (during weighted voting) consists of only 20 templates, so
    the most we could hope for is a "squashing" upwards of the curve towards 
    the 20-template limit (which is the same across all cases in this experiment).
  - we do see a slight "squashing" effect, with XCORR more strongly than NMI. 
    Perhaps in future experiments with larger template libraries, this could be
    explored further.
- Why don't we use STAPLE? What would we expect from STAPLE or other fusion methods? 
  - couldn't get STAPLE to work with our image formats
  - Idea: expect smoothness across range of templates (no even # dips, below)
  - perhaps more sophisticated fusion methods would boost results over majority 
    vote based techniques.
